A Test of the Frequency Independence Assumption of Power System Coefficients used in Geomagnetically Induced Current Estimates

### 2019-07-18 paper.tex
Updates from Overleaf

Motivated by the results described in the following section, in which Model~2 provides a significantly better representation of GIC than Model~1, we have also considered two additional models that can be used to specify GIC. In these two models, a direct measurement of the geoelectric field is not used; instead, the driver is the geomagnetic field, which is much more frequently measured, and as a result, these models may be of practical interest.

@@

Estimates of $a(\omega)$ and $b(\omega)$ are made at evaluation frequencies, $\omega_e \equiv 2\pi f_e$, by performing a ordinary least squares regression (OLS) using Equation~\ref{model2} and a set $G_E(\omega)$ and $\mathbf{E}(\omega)$ in a band of frequencies around $f_e$ from the DFT (discrete Fourier transform) of 24 hours of 1-s cadence data. Following the suggestion of \cite{Simpson2005}, the 28 values of $f_e$ start at $0.25$ Hz and decrease by a factor of $\sqrt{2}$ and end at $1/43200$~Hz; if a $f_e$ value computed in this way is not an integer multiple of $1/86400$, the nearest smaller integer multiple is used. The periods associated with the evaluation frequencies are listed in Table~\ref{evaluationperiods}. The regression for each $f_e$ uses all DFT frequencies with values in a window with a width of twice the separation from the next lower value of $f_e$ and the given $f_e$ with the exceptions that the lowest evaluation frequency band has a lower bound of $1/86400$~Hz and highest evaluation frequency band has an upper bound of $0.375$~Hz.

@@

Several modifications to this basic procedure were considered. Robust regression \citep{Egbert1986} with a Huber weighting constant of $1.345$ and an upper limit of 50 iterations and bi-square weighting with a constant of $4.685$ (both constants provide 95\% efficiency for gaussian-distributed errors), both with no hard-cutoff step; time domain pre-whitening and tapering; and weighting the frequencies around each evaluation frequency using Parzen window weights. When using these additional steps individually, we find that the conclusions made regarding the differences each model's prediction ability do not change because the differences in each model's prediction performance are within the calculated uncertainty of the used method. Transfer function coefficients found using these modifications are generally within the error bars of those for the used procedure when the signal to noise ratio is above $\sim$3 (corresponding to periods longer than $\sim$30~s; At periods below $\sim$30~s, the electric field measurements are significantly influenced by the instrument response characteristics~\citep{Fujii2015}, aliasing, and the computed values are also expected to be significantly biased due to a low signal-to-noise ratio.)

@@

The PE and MSE ratios are shown in Table~\ref{results}. Figure~\ref{predictions} shows the predictions for each model in a 1-day interval selected because the value of the prediction efficiency for each model was in the same order as that shown in Table~\ref{results}, i.e., Model~1 has the lowest PE and Model~4 has the highest. Note that the large errors visible at the start and end of the time intervals are due to the transients discussed in \ref{section:Model_Evaluation} and the first and last 10~minutes of each interval were omitted from metrics calculations.

@@

The primary conclusions of this work follow from the results shown in Table~\ref{results}: the frequency dependent model, Model~2, is significantly better at predicting GIC than the frequency independent model, Model~1; Model~3, which uses an estimate of the electric field based on magnetic field measurements, provides significantly better estimates of GIC than Model~2; Model~4, in which a transfer function that connects GIC to $\mathbf{B}$ was derived directly from the data, provides better estimates of GIC than Model~3.

@@

Below $600$~s (10~min), Model~2 has the largest SN ratio. This result is somewhat visible in the sample interval shown in Figure~\ref{predictions} where the high-frequency fluctuations in the prediction error for Model~2 are the smallest among the models. For these periods, the predictive ability of each model is in the order that may be expected based on the data used to derive the model: Model~2, which uses direct measurements of the electric field, has a larger SN ratio than Model~3, which uses an indirect estimate of the electric field. Model~4 has a slightly larger SN than Model~3, and this is expected because they both use the same input of the magnetic field, but Model~3 has an additional source of uncertainty from the data-estimated coefficients in $\mathbf{Z}$.

Above $600$~s, the ordering of the SN is not consistent with these expectations and is possibly explained by two factors: (1) The measured GIC is based on an integrated electric field along the power line, and the electric field used in Model~2 is a point measurement made 9~km away from the line - it follows that there is an implicit assumption in the model that the electric field is uniform over the length of the line on which GIC is measured; and (2) The error associated with this assumption has less of an influence on the errors in a model that uses the point measurements of the magnetic field. Issues related to this are discussed in the following section.

@@

The average coefficients for Models~2 and~4 are shown in Figure~\ref{Z}--\ref{Phi}. The error bars at each time step or period is the standard error of the 35 values used to compute the average.

@@

Figure~\ref{Z_MT} and~\ref{Phi_MT} show the parameters for $\mathbf{Z}$ in Model~3. Note that electric and magnetic field data from Memanbetsu, Japan were also used by \cite{Fujii2015} to compute $\mathbf{Z}$. When using the \cite{Fujii2015} transfer function for Model~3, the PE was significantly lower (0.58 vs. 0.78). (The \cite{Fujii2015} transfer function was obtained from the IRIS DMC EMTF database [\cite{Kelbert2011}]). Although not critical to the conclusions in this work, a preliminary investigation suggests several explanations. We have found that when using continuous one-month intervals prior to 2006 to compute $\mathbf{Z}$, our calculated transfer function matches the \cite{Fujii2015} transfer function to within the error bars except below periods of 100~s, which is likely explained by the issues with the electric field response in the 1-second-cadence measurements discussed in \cite{Fujii2015}. After 2006, our calculated transfer functions differ significantly from that in \cite{Fujii2015} - they are shifted downward by approximately 10-50\%. In addition, we have found that the estimated transfer function is sensitive to whether or not segment averaging was used. Due to the large gaps between and the varying length of continuous segments of data used in this work, transfer function parameters were computed using one-day segments of data and then averaged. We have found that if a continuous set of measurements over approximately 21 days is used to compute a single transfer function, the transfer function magnitudes are up to 25\% larger than if 21 transfer functions were computed and then averaged. This result is unexpected, and extensive successful verification tests of our MT calculations along with the fact that this effect has observed by both authors using two different MT calculation codes on different data sets (the other code was MT software distributed with MT hardware), suggests the explanation is not due to an error in the code used her to compute transfer functions. We expect to consider this in a future investigation as it suggests that stack averaging can be a significant source of transfer function bias and note that the most likely explanation involves non-stationarity in the noise.

@@

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{figures/map.pdf}
\caption{Locations of measurement sites for GIC, at the Memanbetsu substation, and the geoelectric and geomagnetic field measurement site at Memambetsu Magnetic Observatory. The red line corresponds to the path of the 187 kV power line. The distance from the observatory to the substation is $\sim$9 km; dotted lines are Apex geomagnetic latitudes \cite{Richmond1995} in 2015.}
\label{map}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{figures/plot_model_summary_aobo_histograms-options-1.pdf}
\caption{Distribution of coefficients in Model~1. Each set of coefficients was computed 35 times using one day of 1-second cadence data. The circle/dot on the horizontal axis indicates the average of each histogram: $a_o$ = 73 A/(V/km) and $b_o$ = -72 A/(V/km).}
\label{histogram}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{figures/plot_model_summary_Z_MT-options-1.pdf}
\caption{Frequency domain transfer functions for the coefficients in Model~3.}
\label{Z_MT}
%\end{figure}

@@

%\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{figures/plot_model_summary_Phi_MT-options-1.pdf}
\caption{Frequency domain phase values for the coefficients in Model~3.}
\label{Phi_MT}
\end{figure}

@@

\begin{table}
\caption{Periods associated with evaluation frequency bands used for regression and smoothing spectra. \# is the evaluation frequency number, $N$ is the number of DFT points in the window, and $T_e=1/f_e$. The window range is $T_l$ through $T_h$.}
\centering
\begin{tabular}{l l l l l}
\hline \\
\# & $N$ & $T_e$ [s] & $T_l$ [s] & $T_h$ [s] \\
\hline \\
1 & 2 & 43200 & 28800 & 86400 \\
2 & 2 & 28800 & 21600 & 43200 \\
3 & 3 & 21600 & 14400 & 43200 \\
4 & 4 & 14400 & 9600 & 28800 \\
5 & 5 & 10800 & 7200 & 21600 \\
6 & 6 & 7854.5 & 5400 & 14400 \\
7 & 8 & 5760 & 3927.3 & 10800 \\
8 & 12 & 3927.3 & 2618.2 & 7854.5 \\
9 & 16 & 2880 & 1920 & 5760 \\
10 & 22 & 2009.3 & 1350 & 3927.3 \\
11 & 31 & 1440 & 960 & 2880 \\
12 & 43 & 1016.5 & 680.3 & 2009.3 \\
13 & 61 & 720 & 480 & 1440 \\
14 & 85 & 511.2 & 341.5 & 1016.5 \\
15 & 120 & 361.5 & 241.3 & 720 \\
16 & 170 & 255.6 & 170.4 & 511.2 \\
17 & 240 & 180.8 & 120.5 & 361.5 \\
18 & 338 & 128 & 85.4 & 255.6 \\
19 & 478 & 90.5 & 60.3 & 180.8 \\
20 & 676 & 64 & 42.7 & 128 \\
21 & 956 & 45.2 & 30.2 & 90.5 \\
22 & 1351 & 32 & 21.3 & 64 \\
23 & 1910 & 22.6 & 15.1 & 45.2 \\
24 & 2701 & 16 & 10.7 & 32 \\
25 & 3819 & 11.3 & 7.5 & 22.6 \\
26 & 5401 & 8 & 5.3 & 16 \\
27 & 7638 & 5.7 & 3.8 & 11.3 \\
28 & 10801 & 4 & 2.7 & 8 \\
\hline \\
\end{tabular}
\label{evaluationperiods}
\end{table}

>>>>

Motivated by the results described in the following section, in which Model~2 provides a significantly better representation of GIC than Model~1, we have also considered two additional models that can be used to estimate GIC. In these two models, a direct measurement of the geoelectric field is not used; instead, the driver is the geomagnetic field, which is much more frequently measured, and as a result, these models may be of practical interest.

@@

Estimates of $a(\omega)$ and $b(\omega)$ are made at evaluation frequencies, $\omega_e \equiv 2\pi f_e$, by performing a ordinary least squares regression (OLS) using Equation~\ref{model2} and a set $G_E(\omega)$ and $\mathbf{E}(\omega)$ in a band of frequencies around $f_e$ from the DFT (discrete Fourier transform) of 24 hours of 1-s cadence data. Following the suggestion of \cite{Simpson2005}, the 28 values of $f_e$ start at $0.25$ Hz and decrease by a factor of $\sqrt{2}$ and end at $1/43200$~Hz; if a $f_e$ value computed in this way is not an integer multiple of $1/86400$, the nearest smaller integer multiple is used. The periods associated with the evaluation frequencies are listed in Table~\ref{evaluationperiods}. The regression for each $f_e$ uses all DFT frequencies with values in a band with a width of twice the separation from the next lower value of $f_e$ and the given $f_e$ with the exceptions that the lowest band has a lower bound of $1/86400$~Hz and highest band has an upper bound of $0.375$~Hz.

@@

Several modifications to this basic procedure were considered. Robust regression \citep{Egbert1986} with a Huber weighting constant of $1.345$ and an upper limit of 50 iterations and bi-square weighting with a constant of $4.685$ (both constants provide 95\% efficiency for gaussian-distributed errors), both with no hard-cutoff step; time domain pre-whitening and tapering; and weighting the frequencies around each evaluation frequency using Parzen window weights. When using these additional steps individually, we find that the conclusions made regarding the differences each model's prediction ability do not change because the differences in each model's prediction performance are within the calculated uncertainty of the used method. Transfer function coefficients found using these modifications are generally within the error bars of those for the used procedure when the signal to noise ratio is above $\sim$3 (corresponding to periods longer than $\sim$30~s; At periods shorter than $\sim$30~s, the electric field measurements are significantly influenced by the instrument response characteristics~\citep{Fujii2015}, aliasing, and the computed values are also expected to be significantly biased due to a low signal-to-noise ratio.)

@@

The PE and MSE ratios are shown in Table~\ref{results}. Figure~\ref{predictions} shows the predictions for each model in a 1-day interval selected because the value of the prediction efficiency for each model was in the same order as that shown in Table~\ref{results}, i.e., Model~1 has the lowest PE and Model~4 has the highest. Note that the large errors visible at the start and end of the time intervals are due to the transients discussed in Section~\ref{section:Model_Evaluation} and the first and last 10~minutes of each interval were omitted from metrics calculations.

@@

The primary conclusions of this work follow from the results shown in Table~\ref{results}: the frequency dependent model, Model~2, is significantly better at estimating GIC than the frequency independent model, Model~1; Model~3, which uses an estimate of the electric field based on magnetic field measurements, provides significantly better estimates of GIC than Model~2; Model~4, in which a transfer function that connects GIC to $\mathbf{B}$ was derived directly from the data, provides better estimates of GIC than Model~3.

@@

Below $600$~s, Model~2 has the largest SN ratio. This result is somewhat visible in the sample interval shown in Figure~\ref{predictions} where the high-frequency fluctuations in the prediction error for Model~2 are the smallest among the models. For these periods, the predictive ability of each model is in the order that may be expected based on the data used to derive the model: Model~2, which uses direct measurements of the electric field, has a larger SN ratio than Model~3, which uses an indirect estimate of the electric field. Model~4 has a slightly larger SN than Model~3, and this is expected because they both use the same input of the magnetic field, but Model~3 has an additional source of uncertainty from the data-estimated coefficients in $\mathbf{Z}$.

Above $600$~s, the ordering of the SN is not consistent with these expectations and is possibly explained by two factors: (1) the measured GIC is based on an integrated electric field along the power line, and the electric field used in Model~2 is a point measurement made 9~km away from the line - it follows that there is an implicit assumption in the model that the electric field is uniform over the length of the line on which GIC is measured; and (2) the error associated with this assumption has less of an influence on the errors in a model that uses the point measurements of the magnetic field. Issues related to this are discussed in the following section.

@@

The average coefficients for Models~2 and~4 are shown in Figure~\ref{Z}--\ref{Phi}. The error bars at each period correspond to the standard error of the 35 values used to compute the average.

@@

Figure~\ref{Z_MT} and~\ref{Phi_MT} show the parameters for $\mathbf{Z}$ in Model~3. Note that electric and magnetic field data from Memanbetsu, Japan were also used by \cite{Fujii2015} to compute $\mathbf{Z}$. When using the \cite{Fujii2015} transfer function for Model~3, the PE was significantly lower (0.58 vs. 0.78). (The \cite{Fujii2015} transfer function was obtained from the IRIS DMC EMTF database [\cite{Kelbert2011}]). Although not critical to the conclusions in this work, a preliminary investigation suggests several explanations. We have found that when using continuous one-month intervals prior to 2006 to compute $\mathbf{Z}$, our calculated transfer function matches the \cite{Fujii2015} transfer function to within the error bars except below periods of 100~s, which is likely explained by the issues with the electric field response in the 1-second-cadence measurements discussed in \cite{Fujii2015}. After 2006, our calculated transfer functions differ significantly from that in \cite{Fujii2015} - they are shifted downward by approximately 10-50\%. In addition, we have found that the estimated transfer function is sensitive to whether or not segment averaging was used. Due to the large gaps between and the varying length of continuous segments of data used in this work, transfer function parameters were computed using one-day segments of data and then averaged. We have found that if a continuous set of measurements over approximately 21 days is used to compute a single transfer function, the transfer function magnitudes are up to 25\% larger than if 21 transfer functions were computed and then averaged. This result is unexpected, and extensive successful verification tests of our MT calculations along with the fact that this effect has observed by both authors using two different MT calculation codes on different data sets (the other code used was LEMI MT V1.4), suggests the explanation is not due to an error in the code used here to compute transfer functions. We expect to consider this in a future investigation as it suggests that stack averaging can be a significant source of transfer function bias and note that the most likely explanation involves non-stationarity in the noise.

@@

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{figures/map.pdf}
\caption{Locations of measurement sites for GIC, at the Memanbetsu substation, and the geoelectric and geomagnetic field measurement site at Memambetsu Magnetic Observatory. The red line corresponds to the path of the 187 kV power line. The distance from the observatory to the substation is $\sim$9 km and dotted lines are Apex geomagnetic latitudes \citep{Richmond1995} in 2015.}
\label{map}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{figures/plot_model_summary_aobo_histograms-options-1.pdf}
\caption{Distribution of coefficients in Model~1. Each set of coefficients was computed 35 times using one day of 1-second cadence data. The circle or dot on the horizontal axis indicates the average of a histogram: $a_o$ = 73 A/(V/km) and $b_o$ = -72 A/(V/km).}
\label{histogram}
\end{figure}

@@

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{figures/plot_model_summary_Z_MT-options-1.pdf}
\caption{Frequency domain transfer functions for the $\mathbf{Z}$ coefficients used in Model~3.}
\label{Z_MT}
%\end{figure}

@@

%\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{figures/plot_model_summary_Phi_MT-options-1.pdf}
\caption{Frequency domain phase values for the $\mathbf{Z}$ coefficients used in Model~3.}
\label{Phi_MT}
\end{figure}

@@

\begin{table}
\caption{Periods associated with evaluation frequency bands used for regression and smoothing spectra. \# is the evaluation frequency number, $N$ is the number of DFT points in the band, and $T_e\equv 1/f_e$. The band range is from $T_l$ through $T_h$.}
\centering
\begin{tabular}{l l l l l}
\hline \\
\# & $N$ & $T_e$ [s] & $T_l$ [s] & $T_h$ [s] \\
\hline \\
1 & 2 & 43200 & 28800 & 86400 \\
2 & 2 & 28800 & 21600 & 43200 \\
3 & 3 & 21600 & 14400 & 43200 \\
4 & 4 & 14400 & 9600 & 28800 \\
5 & 5 & 10800 & 7200 & 21600 \\
6 & 6 & 7854.5 & 5400 & 14400 \\
7 & 8 & 5760 & 3927.3 & 10800 \\
8 & 12 & 3927.3 & 2618.2 & 7854.5 \\
9 & 16 & 2880 & 1920 & 5760 \\
10 & 22 & 2009.3 & 1350 & 3927.3 \\
11 & 31 & 1440 & 960 & 2880 \\
12 & 43 & 1016.5 & 680.3 & 2009.3 \\
13 & 61 & 720 & 480 & 1440 \\
14 & 85 & 511.2 & 341.5 & 1016.5 \\
15 & 120 & 361.5 & 241.3 & 720 \\
16 & 170 & 255.6 & 170.4 & 511.2 \\
17 & 240 & 180.8 & 120.5 & 361.5 \\
18 & 338 & 128 & 85.4 & 255.6 \\
19 & 478 & 90.5 & 60.3 & 180.8 \\
20 & 676 & 64 & 42.7 & 128 \\
21 & 956 & 45.2 & 30.2 & 90.5 \\
22 & 1351 & 32 & 21.3 & 64 \\
23 & 1910 & 22.6 & 15.1 & 45.2 \\
24 & 2701 & 16 & 10.7 & 32 \\
25 & 3819 & 11.3 & 7.5 & 22.6 \\
26 & 5401 & 8 & 5.3 & 16 \\
27 & 7638 & 5.7 & 3.8 & 11.3 \\
28 & 10801 & 4 & 2.7 & 8 \\
\hline \\
\end{tabular}
\label{evaluationperiods}
\end{table}
